\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}

\title{A Dream I Had}
\author{Hanif Bin Ariffin} 

\begin{document}

\maketitle

\section{Introduction}

As I am writing this I had just woken up from a very strange dream.
I am very hungry, and as I wait for my food to finish, I thought I had to put this dream into writing.
It was too strange a dream, and I don't often dream too.

The dream was split into 3 parts.
I shall call them the Prologue, a transitioning scene (which I decided to call \textit{The School Affair}) and the Epilogue.

\section{Epilogue}

The dream began by a scene that took place in what looked like a racing track.
Not the usual kind of racing track that's flat; but of hills, sands and rough people.
I can't help but felt like these were rough people.
Perhaps after a day of hard work in the mines, they come to this place in the evenings.
Why are they here?
Aren't they tired?
Regardless, they cheered and shouted throughout the show.
Well, at least for part of it.

Here, I was \textit{reminded} that all these attendees were all killed mysteriously in the stadium.
Perhaps it was only as an aside, but I was too curious.
Isn't it the previlige of the dreamers to dream what they want?

It was then that I was given a firsthand view of the event that killed everyone.
Apparently, one of the racer car went off track and mauled everyone.
It had veered off the track near the finishing line and ran through an ocean of people.
There were no blood, only a mile of dead bodies and up ahead, a girl.
A wolf girl?
I can't quite make it.
She was shining, lights are beaming from her back.
It was quite blinding, but I could clearly make out her nose.
They are felines'.
It was perfectly black too, and maybe thats why I could make it out?

As the car was about to crush her, I \textit{woke} up to a scene of high school students walking to school.
It was in the middle of a desert, with the sun alone in the blue sky; it was cloudless.
It was evidently hot, and yet none seemed to care.

The main character of this scene is evidently a girl, who is seen surrounded by her peers.

\section{The School Affair}

It was quite clear that she was bullied, and yet she does not seem to notice.
Maybe she did notice but not care?
I am not quite sure myself.
But she took the appearance of someone in good company, and yet I felt deeply sorry for her.

As she was harrased, she realized that she had been late for something.
She ran through the crowd and escaped the abuse towards her room.
By the time she arrived, of which was instant to me, it was nighttime.
There she met her roommates.

I couldn't make out the details of this particular scene.
I could only remember the following details.
Her bed was by the window at the other side of the door.
I think on the right, was her roommates'?
On the left side, there's a shelf of books.
The details of the room itself do not matter, but its the only details that I do remember.

Then, she felt again that she's late for something.
``Ah, the meeting with the councillor!'', she said to herself.
She was clearly \textit{very} late that she didn't even prepare.
She ran straight to the hallway, took the steps down a spiral stairs and heads to the office.

There were a great many people there, and she was in a queue.
Before her was her classmates, of which details I will omit here.
Strangely, only she seemed troubled being there.
Everyone else felt happy, smiling and laughing while they wait their turn.
And yet she felt suspended --- why is she being called?

Here, again, we are reminded of the massacre.
Apparently, as the car was about to crash the girl, the light coming out of her grew brighter and brighter.
Then, it exploded; and that was what killed everyone.
As if to remind me, the dreamer, I had a glimpse of the event on the way to school.
Now, it was clear that she was mocked.

Her name was called.
She stood up from the chair she was sitting just infront of the room and began striding there.
However, as she arrived at the door, she heard an extremely loud scream.
It was a screetching of a mad woman.

\section{Epilogue}

After this deafening scream, I was pulled back to the place on their way to school.
Here, she was alone except for another girl.
The girl was tall, slender and have short hair.
The girl was walking ahead of her, but seemed to keep her pace.

Our main character then ran to her, as if to catch up.
The girl reminds our girl that she was bullied, and that she must stood for herself.
But she keeps rejecting the idea that she was bullied.
``They are just like that'', she said.
The girl seemed mad after this, and paced forward as if to lose her.

As she began pacing herself, to match her pace, she began to stumble.
As she was about to fall to her face however, he leg transformed into that of a goat.
It had a very thin foot, but the upper part was extremely buff and the wools were bright red.
Realizing this horror, our girl screamed and screamed, and then fainted.

The other girl, who was very far away now, began running back.
The scene had a strong smell of dense air.
Indeed, a black pool of black oil was growing out of the collapsed body.
As she lifted her face, it was covered in thick oil.

As if materialized out of nowhere, she wiped this thick oil with a piece of cloth she kept handy.
She could wipe everything except a strange rectangular mark on her forehead.
She deduced that it ws the source of the thick oil.
What stranger, to me, the dreamer, was her face.

She was smiling, grinning, stretching her lips forming a deep crescent.
It was the kind of smile when someone realized how stupid one is.
The kind of smile that one makes when coping with grief, as if reminding oneself how much of a joke life is.
Behind this smiling face however, I could make out anger, rage, despise.
I think she now realized that she was indeed bullied.
All this time, all these years, she were abused and now she recalled it all.
As she stood up herself, she was looking forward, and forward was the school.

Then I woke up.

\end{document}
