\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}

%opening
\title{White Nights : \textit{A Sentimental Love Story}}
\author{Fyodor Dostoevsky}

\begin{document}

\maketitle

\begin{abstract}

    And was it his destined part

    Only one moment in life

    To be close to your hearts? \dots

    --- Evan Turgenev

\end{abstract}

\section{First Night}

It was a lovely night, one of those nights, dear reader, which can only happen when you are young.
The sky was so bright and starry that when you looked at it the first question that came into your mind was whether it was really possible that all sorts of bad-tempered and unstable people could live under such a glorious sky.
It is a question, dear reader, that would occur only to a young man, but may the good Lord put it  into your head as often as possible! \dots
The mention of bad-tempered and unstable people reminds me that during the whole of this day my behaviour has been above reproach.
When I woke up in the morning I felt strangely depressed, a feeling I could not shake off for the better part of the day.
All of a sudden it seemed to me as though I, the solitary one, had been forsaken by the whole world, and that the whole world would have nothing to do with me.
You may ask who ``the whole world'' is.
For I am afraid, I have not been lucky in acquiring a single acquaintance in Petersburg during the eight years I have been living there.
But what do I want acquaintances for?
I know the whole of Petersburg without them, and that, indeed, was the reason why it seemed to me that the whole world had forsaken me when the whole town suddenly arose and left for the country.

\end{document}
