\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}

\title{Postmortem: Gas Powered Games' Dungeon Siege}
\author{Bartosz Kijanka}

\begin{document}

\maketitle

\begin{abstract}

    The following text is a transcript of this \href{https://www.gamasutra.com/view/feature/131321/postmortem_gas_powered_games_.php}{article}

\end{abstract}

The development of Dungeon Siege was about much more than the RPG you may have played by now. Dungeon Siege was a Herculean effort by a small group of people who simultaneously started Gas Powered Games, built their first RPG, and made a hit game.

Prior to his career as a prophet, Chris Taylor, Gas Powered Games' president and founder, was responsible for the creation of both the hit RTS game Total Annihilation, as well as Cavedog Entertainment, from which it came. For reasons that are now obscured by the mists of time, Chris was inspired to try his hand at something new, and Gas Powered Games was born. A number of us who worked with Chris in the past, primarily on Total Annihilation, had similar passions, and it wasn't long before we found ourselves cozy once again in a closet office in Kirkland, Wash. We had just finished work on a successful RTS game, a genre we loved working on, but for a number of reasons we opted to try a different genre. We knew we were skilled enough to make another RTS, but the world was swimming in them at that time, so it seemed like a good opportunity to try something new. We all liked the fantasy aesthetic and there were very few good RPG games at that time, so somehow doing an RPG seemed like the natural choice.

We were starting completely from scratch. We had to find an office and buy phones, fax machines, desks, and all those little things most of us take for granted working at an established company. Our first engineering meeting was held at Chris's house with a whiteboard on the floor and a half dozen of us huddled around it. A month later, when we finally did find a space, we had to make it usable. I remember Chris crimping network cables and climbing up on ladders through ceiling tiles running the network wire in the new office. I also clearly remember my first day in that office, when I built my own PC from random parts we bought, unfolding a cheap Costco table on which to set it all up. I loaded up MS Dev and faced an empty header file, which at that moment represented our entire technology base.

Chris repeatedly told us we were in for a wild ride, but I don't think that prophecy sank in until years later. So merrily and naively, we set out on what turned out to be a four-year journey that was as challenging as it was educational.

\end{document}
